Accelerated evolution of convective simulations

### 2018-02-21 tex/bvp_paper.tex
some text cleaning

\section{Introduction}
\label{sec:intro}
Natural convection occurs in the presence of disparate timescales which
prohibit numericists from studying realistic models of natural systems.  For example,
flows in the convection zone of stars like the Sun are often at low Mach number
(Ma) deep in the interior.
Explicit methods which are bound by the Courant-Friedrich-Lewy
(CFL) timestep limit must resolve the fastest motions (sound
waves), resulting in timesteps which are prohibitively
small for studies of the deep, low-Ma motions. These systems are numerically
stiff, and the disparity between
the sound crossing time and the convective flows have made studies of low-Ma stellar
convection difficult. Traditionally, approximations such as
the anelastic approximation, in which sound waves are explicitly filtered out,
have been used to study low-Ma flows \cite{brown&all2010, featherstone&hindman2016}.
More recently, advanced numerical techniques which use implicit or mixed
implicit-explicit timestepping mechanisms have made it feasible to study
convection at low Mach numbers \cite{viallet&all2011, viallet&all2013, viallet&all2016, lecoanet&all2014,
anders&brown2017, bordwell&all2018}, and careful studies of deep convection which
would have been impossible a decade ago are now widely accessible.

While solutions to problem of disparate \emph{dynamical} timescales have proven useful,
the difference between convective timescales and system relaxation timescales remains
a significant problem facing studies of convection. 
As modern simulations aim to model natural convection
by increasing into the high-Rayleigh-number (Ra) regime,
the ratio of the thermal diffusion timescale to dynamical timescales
grows \cite{anders&brown2017}.
Solar convection is a prime example of this phenomenon.
Dynamical timescales in the solar convective zone are relatively short 
(10 min overturn at solar surface, one month solar
rotation rate) compared to the Kelvin-Helmholtz timescale of
$3 \cdot 10^7$ years \cite{stix2003}.  
Furthermore, as dynamical and diffusive timescales separate, 
simulations become more turbulent. Increasingly turbulent motions 
require finer grid meshes and smaller timesteps
to capture advective dynamics. Thus, the progression of simulations into the high-Ra
regime of natural convection is slowed by two simultaneous effects: timestepping
through a single convective overturn time becomes more computationally expensive
while the number of overturn times required for systems to reach thermal equilibration
grows.

The vast difference between convective and diffusive timescales has long plagued
numericists studying convection, and a plethora of approaches has been employed to
study thermally converged solutions. One popular method for accelerating the convergence
of high-Ra solutions is by ``Bootstrapping,'' or the process of using the flow
fields in a converged solution at lower Ra as initial conditions for a simulation at a higher
Ra.  This method has been used with great success \cite{johnston&doering2009, verzicco&camussi1997},
but it is not without its faults.  Bootstrapped solutions are succeptible to hysteresis
effects, in which large-scale convective structures present in the
low Ra solution imprint onto the dynamics of the new, high Ra solution. 
Another commonly-used tactic in
moderate-Ra simulations is to use 
a simple model of the full convective state as initial conditions.  
For example, past studies have used a linear eigenvalue solve to set the initial
convective state \cite{hurlburt&all1984} or used an axisymmetric solution 
as initial conditions for convection in a 3D cylinder \cite{verzicco&camussi1997}. 
In certain systems, 
the approximate state of the evolved solution can be estimated. There, an
appropriate set of initial conditions can either be solved for analytically
\cite{couston&all2017} or by using knowledge of Mixing Length Theory or other convective
theories to adjust the initial profile towards the proper adiabatic state \cite{brandenburg&all2005}.

Despite the plethora of methods that have been used,
the most straightforward way to achieve a thermally converged solution
is to evolve a convective simulation through a thermal timescale. Some modern
studies do just that \cite{featherstone&hindman2016}, but it is becoming
increasingly difficult. Such evolution is
\emph{expensive}, and state-of-the-art simulations at the highest values of Ra
can only be reasonably run
for tens to hundreds of buoyancy times \cite{stevens&all2011}, much less the
thousands of buoyancy timescales required for thermal convergence.

@@

@@

@@

@@

\section{Extensions of the AE method}
\label{sec:extensions}
Here we have used \RB convection as a test case for the AE method, but we argue that
the true power of this method is in its extensions to more complicated studies.
Such extensions are beyond the scope of this work, but we will briefly describe
avenues in which extensions to the AE method will be useful.
For example, in studies of stratified convection where the thermal diffusivity
is inversely proportional to the density \cite{anders&brown2017}, AE solutions
will allow for the careful study of a broader range of parameter space. 
Further, in studies where stable and convecting regions are paired
\cite{hurlburt&all1986, couston&all2017}, the stable layers cannot rely on convective
motions to restratify them and instead evolve on the order of the slow thermal
timescale. In
fully-convective domains, such as those studied here, the atmosphere evolves
much more rapidly than the thermal diffusion time (see e.g., Fig. \ref{fig:time_trace}a).

We note also that a further strength of the AE method is its generality.  One
need only derive the steady-state, horizontally-averaged equations for a given
equation set (e.g., Eqns. \ref{eqn:bouss_BVP_momentum} \& \ref{eqn:bouss_BVP_energy})
and couple that information with knowledge of the current dynamics (Eqn. \ref{eqn:bvp_ratios})
and the boundary conditions. Thus this method can be easily applied to more complex
systems, and merely require that the experimenter understand the energy transport
present in their systems.

We have only studied solutions in which we use mixed thermal boundary conditions,
as in Eqn. \ref{eqn:bcs}. However, this method can be easily applied to fixed-temperature
boundary conditions, as well.  Rather than knowing the total final flux (F$_{\text{bot}}$)
through the system a priori, it must be measured using the ratios in
Eqn. \ref{eqn:bvp_ratios}.  We hesitate to use the AE method for systems in which
both boundaries fix the flux, as such systems have degenerate solutions for the temperature.

\section{Conclusion}
\label{sec:conclusion}
Here we have described a method for converging convective simulations using
a method of Accelerated Evolution (AE), and compared it to Standard
Evolution (SE) achieved by running simulations for a thermal timescale.
We have shown that the AE method produces dynamics that are similar to SE
dynamics on much shorter timescales, and have discussed differences that arise
between the two methods.

@@

\section{Table of Runs}
\label{appendix:run_table}
\begin{center}
\begin{tabularx}{\textwidth}{ X X X X X X X X X }
\hline															
$S$	&	Ra	&	nz	&	nx, ny	&	$t_{\text{therm}}$	&	$t_{\text{avg}}$	&	Nu$_{\text{SE}}$	&	Nu$_{\text{AE}}$	\\
\hline															
$10^{1/3}$	&	$2.79 \cdot 10^3$	&	32	&	64	&	$52.8$	&	100	&	1.46	&	1.46	\\
$10^{2/3}$	&	$6.01 \cdot 10^3$	&	32	&	64	&	$77.6$	&	100	&	1.95	&	1.95	\\
$10^1$	&	$1.30 \cdot 10^4$	&	32	&	64	&	$114$	&	100	&	2.43	&	2.42	\\
$10^{1 + 1/3}$	&	$2.79 \cdot 10^4$	&	32	&	64	&	$167$	&	100	&	2.54	&	2.54	\\
$10^{1 + 2/3}$	&	$6.01 \cdot 10^4$	&	32	&	64	&	$245$	&	100	&	3.14	&	3.14	\\
$10^2$	&	$1.30 \cdot 10^5$	&	64	&	128	&	$360$	&	100	&	3.8	&	3.8	\\
$10^{2 + 1/3}$	&	$2.79 \cdot 10^5$	&	64	&	128	&	$528$	&	100	&	4.71	&	4.71	\\
$10^{2 + 2/3}$	&	$6.01 \cdot 10^5$	&	64	&	128	&	$776$	&	100	&	5.5	&	5.5	\\
$10^3$	&	$1.30 \cdot 10^6$	&	128	&	256	&	$1.14 \cdot 10^3$	&	200	&	6.4	&	6.33	\\
$10^{3 + 1/3}$	&	$2.79 \cdot 10^6$	&	128	&	256	&	$1.67 \cdot 10^3$	&	500	&	6.87	&	6.95	\\
$10^{3 + 2/3}$	&	$6.01 \cdot 10^6$	&	256	&	512	&	$2.45 \cdot 10^3$	&	500	&	7.54	&	7.59	\\
$10^4$	&	$1.30 \cdot 10^7$	&	256	&	512	&	$3.60 \cdot 10^3$	&	500	&	8.83	&	8.83	\\
$10^{4 + 1/3}$	&	$2.79 \cdot 10^7$	&	256	&	512	&	$5.28 \cdot 10^3$	&	500	&	10.13	&	10.14	\\
$10^{4 + 2/3}$	&	$6.01 \cdot 10^7$	&	256	&	512	&	$7.76 \cdot 10^3$	&	500	&	11.65	&	11.69	\\
$10^5$	&	$1.30 \cdot 10^8$	&	512	&	1024	&	$1.14 \cdot 10^4$	&	500	&	14.02	&	14.18	\\
$10^{5 + 1/3}$	&	$2.79 \cdot 10^8$	&	512	&	1024	&	$1.67 \cdot 10^4$	&	500	&	N/A	&	16.21	\\
$10^{5 + 2/3}$	&	$6.01 \cdot 10^8$	&	512	&	1024	&	$2.45 \cdot 10^4$	&	500	&	N/A	&	18.58	\\
$10^6$	&	$1.30 \cdot 10^9$	&	1024	&	2048	&	$3.60 \cdot 10^4$	&	500	&	N/A	&	22.132	\\
$10^7$	&	$1.30 \cdot 10^{10}$	&	2048	&	4096	&	$1.14 \cdot 10^5$	&	200	&	N/A	&	TBD	\\
\\ \hline \hline								\\							
$10^1$	&	$1.30 \cdot 10^4$	&	32	&	64,64	&	$114$	&	100	&	2.42	&	2.42	\\
$10^2$	&	$1.30 \cdot 10^5$	&	64	&	128,128	&	$360$	&	100	&	3.97	&	4	\\
$10^3$	&	$1.30 \cdot 10^6$	&	128	&	256,256	&	$1.14 \cdot 10^3$	&	500	&	6.27	&	6.27	\\
$10^4$	&	$1.30 \cdot 10^7$	&	256	&	512,512	&	$3.60 \cdot 10^3$	&	500	&	9.92	&	9.88	\\
\hline															
\end{tabularx}
\end{center}

>>>>

%%%%%%%%%%%%
%%%%%%%%%%%
% INTRO
%%%%%%%%%%%
%%%%%%%%%%%%

\section{Introduction}
\label{sec:intro}
Natural convection occurs in the presence of disparate timescales which
prohibit numericists from studying realistic models of natural systems.  For example,
flows in the convection zone of stars like the Sun are characteristically low Mach number
(Ma) in the deep interior.
Explicit timestepping methods which are bound by the Courant-Friedrich-Lewy
(CFL) timestep limit must resolve the fastest motions (sound
waves), resulting in timesteps which are prohibitively
small for studies of the deep, low-Ma motions. These systems are numerically
stiff, and the difference between
the sound crossing time and the convective overturn time have made studies of low-Ma stellar
convection difficult. Traditionally, approximations such as
the anelastic approximation, in which sound waves are explicitly filtered out,
have been used to study low-Ma flows \cite{brown&all2010, featherstone&hindman2016}.
More recently, advanced numerical techniques which use implicit or mixed
implicit-explicit timestepping mechanisms have made it feasible to study
convection at low Mach numbers \cite{viallet&all2011, viallet&all2013, viallet&all2016, lecoanet&all2014,
anders&brown2017, bordwell&all2018}, and careful studies of deep convection which
would have been impossible a decade ago are now widely accessible.

While solutions to the problem of divergent \emph{dynamical} timescales have proven useful,
the difference between convective timescales and thermal timescales, which
characterize system relaxation, remain a significant problem facing studies of convection. 
Solar convection is a prime example of this phenomenon;
dynamical timescales in the solar convective zone are relatively short 
(10 min overturn at solar surface, one month solar
rotation rate) compared to the Kelvin-Helmholtz timescale of
$3 \cdot 10^7$ years \cite{stix2003}.  
In such a system, it is impossible to resolve the convective dynamics while also
evolving the thermal structure of the system in a meaningful fashion.
As modern simulations aim to model natural convection
by increasing into the high-Rayleigh-number (Ra) regime,
the ratio of the thermal diffusion timescale to dynamical timescales
grows \cite{anders&brown2017}.
Furthermore, as dynamical and thermal timescales separate, 
simulations become more turbulent. Increasingly turbulent motions 
require finer grid meshes and smaller timesteps
to capture advective dynamics. Thus, the progression of simulations into the high-Ra
regime of natural convection is slowed by two simultaneous effects: timestepping
through a single convective overturn time becomes more computationally expensive
while the number of overturn times required for systems to reach thermal equilibration
grows.

The vast difference between convective and thermal timescales has long plagued
numericists studying convection, and an abundance of approaches has been employed to
study thermally converged solutions. One popular method for accelerating the convergence
of high-Ra solutions is by ``bootstrapping'' -- the process of using the flow
fields in a converged solution at lower Ra as initial conditions for a simulation at a higher
Ra.  This method has been used with great success \cite{johnston&doering2009, verzicco&camussi1997},
but it is not without its faults.  Bootstrapped solutions are succeptible to hysteresis
effects, in which large-scale convective structures present in the
low Ra solution imprint onto the dynamics of the new, high Ra solution. 
Another commonly-used tactic in
moderate-Ra simulations is to use 
a simple model of the full convective state as initial conditions.  
For example, past studies have used a linear eigenvalue solve to set the initial
convective state \cite{hurlburt&all1984} or used an axisymmetric solution 
as initial conditions for convection in a 3D cylinder \cite{verzicco&camussi1997}. 
In certain systems, 
the approximate state of the evolved solution can be estimated. There, an
appropriate set of initial conditions can either be solved for analytically
\cite{couston&all2017} or by using knowledge of Mixing Length Theory or other convective
theories to adjust the initial profile towards the proper adiabatic state \cite{brandenburg&all2005}.

Despite the numerous methods that have been used,
the most straightforward way to achieve a thermally converged solution
is to evolve a convective simulation through a thermal timescale. Some modern
studies do just that \cite{featherstone&hindman2016}, but it is becoming
increasingly difficult. Such evolution is
\emph{expensive}, and state-of-the-art simulations at the highest values of Ra
can only be reasonably run
for tens to hundreds of buoyancy times \cite{stevens&all2011}, much less the
thousands of buoyancy timescales required for thermal convergence.

@@

%%%%%%%%%%%%
%%%%%%%%%%%
% EXPERIMENT
%%%%%%%%%%%
%%%%%%%%%%%%

@@

%%%%%%%%%%%%
%%%%%%%%%%%
% AE METHOD
%%%%%%%%%%%
%%%%%%%%%%%%

@@

%%%%%%%%%%%%
%%%%%%%%%%%
% RESULTS
%%%%%%%%%%%
%%%%%%%%%%%%

@@

%%%%%%%%%%%%
%%%%%%%%%%%
% CONCLUSION
%%%%%%%%%%%
%%%%%%%%%%%%

\section{Extensions \& Conclusion}
\label{sec:extensions}
In this work we have used \RB convection as a test case for the AE method, but we argue that
the true power of this method is in its extensions to more complicated studies.
Such extensions are beyond the scope of this paper, but we will briefly describe
areas in which extensions to the AE method should be explored.
We note that a strength of the AE method is its generality.  One
need only derive the steady-state, horizontally-averaged equations for a given
equation set (e.g., Eqns. (\ref{eqn:bouss_BVP_momentum}) \& (\ref{eqn:bouss_BVP_energy}))
and couple that information with knowledge of the current dynamics (Eqn. (\ref{eqn:bvp_ratios}))
and the boundary conditions. Thus this method can be easily applied to more complex
systems, so long as the experimenter understand the energy transport
present in their systems.

In studies of stratified convection where the thermal diffusivity
is inversely proportional to the density \cite{anders&brown2017}, the
thermal timescale deep in the atmosphere is much larger than the timescale
near the surface.  Thus, the AE method can be a very important tool for studying
highly stratified systems which aim to model realistic astrophysical atmospheres.
Through carefully examining the flux in excess of the flux conducted along the
adiabatic temperature gradient, the fractions of Eqn. (\ref{eqn:bvp_ratios})
can be constructed for such a system.  In addition to solving for the average
hydrostatic balance and thermal equilibrium, it is essential to simultaneously
ensure that mass is conserved in truly stratified systems by adding a 
mass-conservation equation, similar to those used in stellar structure models
\cite{paxton&all2011}.

It is common for studies to pair stable and convecting regions
\cite{hurlburt&all1986, brandenburg&all2005, couston&all2017}.
In such studies, particularly when the interface between the stable region
and the convecting region is stiff and motions do not penetrate from the convecting
region into the stable region, the stable region cannot rely on convective
motions to assist in the restratification of the thermal profile.
In fully-convective domains, such as those studied here, the atmosphere evolves
much more rapidly than the thermal diffusion time (see e.g., Fig. \ref{fig:time_trace}a,
where the SE solution is essentially fully converged after $4\cdot 10^3$ frefall units
despite the thermal timescale being roughly $10^4$ frefall units).

We have only studied solutions in which we use mixed thermal boundary conditions,
as in Eqn. \ref{eqn:bcs}. However, this method can be easily applied to fixed-temperature
boundary conditions, as well.  Rather than knowing the total final flux (F$_{\text{bot}}$)
through the system a priori, it must be measured using the ratios in
Eqn. \ref{eqn:bvp_ratios}.  The integrated conductive flux is known in these systems exactly thanks 
to the fixed temperature jump across the domain.  The total flux through the system
($F_{\text{bot}}$) can is thus $\mathcal{P} / \int f_{\text{cond}} dz$ (double-check).
We hesitate to use the AE method for systems in which
both boundaries fix the flux, as such systems have degenerate solutions for the temperature.

In conclusion, here we have described a method for converging convective simulations using
a method of Accelerated Evolution (AE), and compared it to Standard
Evolution (SE) achieved by running simulations for a thermal timescale.
We have shown that the AE method produces dynamics that are similar to SE
dynamics on much shorter timescales, and have discussed differences that arise
between the two methods.

@@

\section{Table of Runs}
\label{appendix:run_table}
\begin{center}
\begin{tabularx}{\textwidth}{ X X X X X X X X X }
\hline															
$S$	&	Ra	&	nz	&	nx, ny	&	$t_{\text{therm}}$	&	$t_{\text{avg}}$	&	Nu$_{\text{SE}}$	&	Nu$_{\text{AE}}$	\\
\hline															
$10^{1/3}$	&	$2.79 \cdot 10^3$	&	32	&	64	&	$52.8$	&	100	&	1.46	&	1.46	\\
$10^{2/3}$	&	$6.01 \cdot 10^3$	&	32	&	64	&	$77.6$	&	100	&	1.95	&	1.95	\\
$10^1$	&	$1.30 \cdot 10^4$	&	32	&	64	&	$114$	&	100	&	2.43	&	2.42	\\
$10^{1 + 1/3}$	&	$2.79 \cdot 10^4$	&	32	&	64	&	$167$	&	100	&	2.54	&	2.54	\\
$10^{1 + 2/3}$	&	$6.01 \cdot 10^4$	&	32	&	64	&	$245$	&	100	&	3.14	&	3.14	\\
$10^2$	&	$1.30 \cdot 10^5$	&	64	&	128	&	$360$	&	100	&	3.8	&	3.8	\\
$10^{2 + 1/3}$	&	$2.79 \cdot 10^5$	&	64	&	128	&	$528$	&	100	&	4.71	&	4.71	\\
$10^{2 + 2/3}$	&	$6.01 \cdot 10^5$	&	64	&	128	&	$776$	&	100	&	5.5	&	5.5	\\
$10^3$	&	$1.30 \cdot 10^6$	&	128	&	256	&	$1.14 \cdot 10^3$	&	200	&	6.4	&	6.33	\\
$10^{3 + 1/3}$	&	$2.79 \cdot 10^6$	&	128	&	256	&	$1.67 \cdot 10^3$	&	500	&	6.87	&	6.95	\\
$10^{3 + 2/3}$	&	$6.01 \cdot 10^6$	&	256	&	512	&	$2.45 \cdot 10^3$	&	500	&	7.54	&	7.59	\\
$10^4$	&	$1.30 \cdot 10^7$	&	256	&	512	&	$3.60 \cdot 10^3$	&	500	&	8.83	&	8.83	\\
$10^{4 + 1/3}$	&	$2.79 \cdot 10^7$	&	256	&	512	&	$5.28 \cdot 10^3$	&	500	&	10.13	&	10.14	\\
$10^{4 + 2/3}$	&	$6.01 \cdot 10^7$	&	256	&	512	&	$7.76 \cdot 10^3$	&	500	&	11.65	&	11.69	\\
$10^5$	&	$1.30 \cdot 10^8$	&	512	&	1024	&	$1.14 \cdot 10^4$	&	500	&	14.02	&	14.18	\\
$10^{5 + 1/3}$	&	$2.79 \cdot 10^8$	&	512	&	1024	&	$1.67 \cdot 10^4$	&	500	&	--	&	16.21	\\
$10^{5 + 2/3}$	&	$6.01 \cdot 10^8$	&	512	&	1024	&	$2.45 \cdot 10^4$	&	500	&	--	&	18.58	\\
$10^6$	&	$1.30 \cdot 10^9$	&	1024	&	2048	&	$3.60 \cdot 10^4$	&	500	&	--	&	22.13	\\
$10^7$	&	$1.30 \cdot 10^{10}$	&	2048	&	4096	&	$1.14 \cdot 10^5$	&	200	&	--	&	40.78	\\
\\ \hline \hline \\															
$10^1$	&	$1.30 \cdot 10^4$	&	32	&	64$\times$64	&	$114$	&	100	&	2.42	&	2.42	\\
$10^2$	&	$1.30 \cdot 10^5$	&	64	&	128$\times$128	&	$360$	&	100	&	3.97	&	4	\\
$10^3$	&	$1.30 \cdot 10^6$	&	128	&	256$\times$256	&	$1.14 \cdot 10^3$	&	500	&	6.27	&	6.27	\\
$10^4$	&	$1.30 \cdot 10^7$	&	256	&	512$\times$512	&	$3.60 \cdot 10^3$	&	500	&	9.92	&	9.88	\\
\hline															
\end{tabularx}
\end{center}
